\textbf{The Illusion Game: A Novel Experimental Paradigm to study Visual Illusions}

### 2022-09-01 manuscript.Rmd

<!-- Target, strength, difficulty -->

@@

<!-- Paradigm -->

This thereby opens the door to a standardizable experimental task in which participants can be asked to make perceptual judgments about the targets (e.g., which segment in the longest) under different conditions of objective difficulty and illusion strength. Moreover, the illusory effect can be set to either be "incongruent" (increasing the difficulty of the task by biasing the perception in the opposite way) or "congruent" (making the task easier). Although visual illusions are inherently tied to subjective perception, this framework radically reverses the traditional paradigm in favor of an approach that could quantify the "objective" effect of illusions by measuring its behavioral effect (error rate and reaction times) on the performance in a perceptual task.

<!-- Goals -->

@@

We generated 56 stimuli for each of the 10 illusion types. These stimuli resulted from the combination of 8 linearly-spread levels levels of task difficulty (e.g., [1, 2, 3, 4, 5, 6, 7], where values correspond to the objective difference between targets, i.e., 1 corresponds to the smallest objective difference and thereby the highest level of difficulty in the task) and 7 levels of illusion strength (3 values (i.e, degrees) of strength on the side of illusion congruence, 3 on the side of illusion incongruence, and 0; e.g., [-3, -2, -1, 0, 1, 2, 3]. Notably, the magnitude of these values correspond to illusion strength (greater values represent greater strength) whereas the valence indicates the illusion congruence (negative sign represents an illusion congruent effect).

The 10 types of illusion stimuli were randomly presented in distinct blocks, within which the 56 stimuli were also displayed in a randomized order. After the first series of 10 blocks was presented, another series was displayed (with the same set of stimuli but with new randomized block and trial orders). All in all, each participant saw 56 different trials per illusion type (10 blocks in total), in 2 series of blocks (amounting to a total of 1120 trials). In each illusion trial, participants were required to evaluate the targets comprising the illusions without being distracted by its biasing contextual information. For example, in a Müller-Lyer illusion trial, participants were instructed to evaluate the length of line segments without being distracted by the direction in which the arrows at its ends are pointing towards. Importantly, participants were told to respond "as fast as possible without making errors" before each block of trials (i.e., an explicit double constraint incorporated to mitigate the inter-individual variability in the speed-accuracy trade off in responses). This task was implemented using *jsPsych* [@deleeuw2014]. For specific details of the instructions given for each illusion type, please refer to the experiment code.

@@

The best models were $log(diff)*strength$ for Delboeuf; $sqrt(diff)*strength$ for Ebbinghaus; $log(diff)*log(strength)$ for Rod and Frame; $sqrt(diff)*sqrt(strength)$ for Vertical-Horizontal; .... For all of the specified models, the effect of illusion strength, task difficulty and their interaction were significant.

@@

<!-- This is likely to be caused by the outline. It will be fixed in the next release of the Pyllusion software. -->

Future studies should investigate reaction times to find the most appropriate models, and / or use models that integrate errors and reaction time (e.g., drift diffusion models).

>>>>

<!-- Target, strength, difficulty -->
Indeed, many visual illusions can be seen as made of *targets* (e.g., same-length lines), which perception is biased by the *context* (e.g., in the Müller-Lyer illusion, the segments appear of different lengths when they end with inwards or outwards pointing arrows). While most of the paradigms used in the literature prompt the subjective extent the context biases the perception (e.g."how much doe the line appear different"; **REFS**), *Pyllusions* allows to create illusions in which the targets are actually more or less different (e.g., one segment is truly longer than the other), and in which the illusion varies in strength (the arrows are more or less pointy).

@@

<!-- Paradigm -->
This opens the door of an experimental task in which participants have to make perceptual judgments about the targets (e.g., which segment in the longest) under different conditions of objective difficulty and illusion strength. Moreover, the illusion effect can be either "incongruent" (making the task even harder by biasing the perception in the opposite way) or "congruent" (making the task easier). Although visual illusions are inherently tied to subjective perception, this framework allows a reversal of the traditional paradigm that could quantify the "objective" effect of illusions by measuring its behavioral effect (error rate and reaction times) on the performance in a perceptual task.

<!-- Goals -->
In the present set of preregistered studies, we will first test this novel paradigm by investigate if the effect of illusion and task difficulty can be manipulated continuously and separated statistically. Then, we will use the paradigm to assess whether 10 different historical illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) share a common latent factor (a long-standing debate). Finally, we will investigate how the the inter-individual sensitivity to illusions relates to other characteristics, such as demographic variables and personality.

@@

We generated 56 stimuli for each of the 10 illusion types. These stimuli resulted from the combination of 8 linearly-spread levels levels of task difficulty (e.g., [1, 2, 3, 4, 5, 6, 7], where 1 corresponds to the higher difficulty - i.e., the smallest objective difference between targets) and 7 levels of illusion strength (3 values of strength on the congruent side, 3 on the incongruent side, and 0; e.g., [-3, -2, -1, 0, 1, 2, 3], where the negative values correspond to congruent illusion strengths).

The 10 illusion blocks were randomly presented, and the order of the 56 stimuli within the block was also randomized. After the first series of 10 blocks, another series was done (with new randomized blocks and trials order). In total, each participant saw 56 different trials per 10 illusion type, repeated 2 times (total = 1120 trials), to which they had to respond "as fast as possible without making errors" (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off). The task was implemented using *jsPsych* (*REF*). See instructions for each illusion type in the experiment code.

@@

The best models were $log(diff)*strength$ for Delboeuf; $sqrt(diff)*strength$ for Ebbinghaus; $log(diff)*log(strength)$ for Rod and Frame; $sqrt(diff)*sqrt(strength)$ for Vertical-Horizontal; .... In all of these models, the effect of illusion strength, task difficulty and their interaction were significant.

For errors, most of the models closely matched their GAMs counterpart, with the exception of Delboeuf for which the GAM suggested a non-monotonic effect of illusion strength with a local minimum at 0.

However, for reaction times, the GAMs suggested a consistent non-linear relationship between the mean RT and illusion strength: as the illusion strength gets very strong, the reaction times are faster. While this is is not surprising (strong illusions are so biasing that it is "easier", i.e., faster, to make the wrong decision), the linear models were not designed to capture this pattern and hence are not good representatives of the underlying dynamics. Because the random-slopes models for RT did not appear to capture well the underlying, likely quadratic, relationship, we decided not to use add them for the individual scores analysis.
<!-- Additionally, it would have increased the parameter space from 20 to 40 variables, which was deemed as statistically unreasonable given our sample size). -->

Though imperfect, we believe that these models capture inter-individual differences more accurately than basic empirical scores (e.g., the total amount of errors, or the average RT). Thus, for each illusion, we have extracted the effect of illusion strength and its interaction with task difficulty (when the illusion incongruent) within each participants. These twenty participant-level scores were subjected to exploratory factor analysis (EFA), which suggested that ...

@@

<!-- This is likely to be caused by the outline. It will be fixed in the next release of the Pyllusion software. -->
Future studies should investigate reaction times to find the most appropriate models, and / or use models that integrate errors and reaction time (e.g., drift diffusion models).
